---
title: "Machine Learning Project"
author: "T. Tilli"
date: "Sunday, January 18, 2015"
output: html_document
---

# Practical Machine Learning:  Peer Assessment 

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 


## Data

The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment. 


## Reproduceablity

In order to reproduce the results, a certrain set of packages is required, as well as setting a pseudo random seed.

```{r}
library(caret)
```

```{r}
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
```
```{r}
library(randomForest)
```
Finally, set the  seed with the following line of code:

```{r}
set.seed(4567)
```

## Loading the data
Loading the training data:

```{r}
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
str(training)
```
There are a lot of variables with NAs.
Loading the testing data:

```{r}
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```

## Partioning the training data 
Partioning set into a dataset for training and one for testing, 60% for myTraining, 40% for myTesting:

```{r}
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)

myTraining <- training[inTrain, ]; 
myTesting <- training[-inTrain, ]

dim(myTraining); 
dim(myTesting)
```

## Cleaning the data

The following transformation steps are used to clean the data:

1.  Cleaning near zero variance variables. First find this variables:
```{r}
trainNZV <- nearZeroVar(myTraining, saveMetrics=TRUE)
```

2. Create a subset of the training data without NZV variables:
```{r}
theNZVvars <- names(myTraining) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt",
"kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
"max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
"var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
"stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
"kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
"max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
"kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell",
"skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell",
"amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
"skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
"max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
"amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
"avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
"stddev_yaw_forearm", "var_yaw_forearm")
myTraining <- myTraining[!theNZVvars]
dim(myTraining)

```

3. Eliminate the dataset ID:
```{r}
myTraining <- myTraining[c(-1)]
```

4. Eliminate the Variables with more than 60% of NAs:

```{r}
train2<- myTraining 
for(i in 1:length(myTraining))
  if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .6 ) {
          
      for(j in 1:length(train2)) {
        if( length( grep(names(myTraining[i]), names(train2)[j]) ) ==1)   {
            train2 <- train2[ , -j] 
        } 
  }
}

dim(train2)

myTraining <- train2
rm(train2)
```

Repeat this steps for the myTesting and testing data sets:
```{r}
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58]) #already with classe column removed
myTesting <- myTesting[clean1]
testing <- testing[clean2]

dim(myTesting)
dim(testing)
```

Coerce the data into the same type in the training and testing data sets:

```{r}
for (i in 1:length(testing) ) {
        for(j in 1:length(myTraining)) {
        if( length( grep(names(myTraining[i]), names(testing)[j]) ) ==1)  {
            class(testing[j]) <- class(myTraining[i])
        }      
    }      
}
testing <- rbind(myTraining[2, -58] , testing) 
testing <- testing[-1,]
```

## ML-Algorithmus I: Decision Tree
```{r}
modFit1 <- rpart(classe ~ ., data=myTraining, method="class")
```


Prediction on myTesting data set:

```{r}
predictions1 <- predict(modFit1, myTesting, type = "class")
```

Calculate Confusing matrix:

```{r}
confusionMatrix(predictions1, myTesting$classe)
```

## ML-Algorithmus II: Random Forest

```{r}
modFit2 <- randomForest(classe ~. , data=myTraining)
```


Prediction on myTesting data set:

```{r}
predictions2 <- predict(modFit2, myTesting, type = "class")
```

Calculate Confusing matrix:

```{r}
confusionMatrix(predictions2, myTesting$classe)
```

Random forest yields much better results. Therefore we will use the random forest model for the test set.

## Generating the files for the Course Project: Submission

Predicting using random forest model on the test set:
```{r}
predictionsTest <- predict(modFit2, testing, type = "class")

predictionsTest
```

Function to generate files with predictions to submit for assignment

```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
```

Write the results to the files:
```{r}
pml_write_files(predictionsTest)
```

## Assessment of Final Predictions
The predictions above were submitted to for automated grading, and all were found to be correct.
